Analysis of NLA17 Dissolved Gas Indicator

### 2020-06-01 scripts/dgIndicatorAnalysis.Rmd
I added two code chunks, Correlogram:  ncf and pgmirness packages and Correlograms by Ecoregions.  The first code chunk compares two packages to make correlograms, and I chose ncf.  I also did some outlier analysis.  The second chunk makes correlograms by the different ecoregions.

```{r results='hide', message=FALSE, warning=FALSE}
# load libraries
library(sf) # spatial data
library(gstat) # lagged scatterplot, variogram
library(tidyverse) # dplyr, ggplot
library(janitor) # clean names
library(spdep) # k-NN and Moran plot

@@

>>>>

```{r results='hide', message=FALSE, warning=FALSE}
# load libraries
library(sf) # spatial data
library(gstat) # lagged scatterplot, variogram
library(tidyverse) # dplyr, ggplot
library(janitor) # clean names
library(spdep) # k-NN and Moran plot
library(ncf) # ncf::correlogram

@@

## Correlograms: ncf and pgmirness packages
```{r, correlogram}

# Much of code below is based on example comparing correlograms among packages ncf, pgmirness, and spdep described in this chapter
# https://link.springer.com/chapter/10.1007/978-3-030-01989-1_5

coords <- data.frame(longitude = dg_1$map_lon_dd, latitude = dg_1$map_lat_dd)

dg_sf1 <- st_as_sf(dg_1, coords = c("map_lon_dd", "map_lat_dd"), 
                  crs = 4269) %>% # standard for lat/lon
  st_transform(5070) # project to CONUS ALBERS for plotting

st_crs(dg_sf1)
names(dg_sf1)
nap_visit1 <- dg_sf1 %>% filter(wsa9 == "NAP" & visit_no == 1)
class(nap_visit1)
st_crs(nap_visit1)

nap_sp_visit1 <- as(nap_visit1, "Spatial")
class(nap_sp_visit1)

# trying ncf package which uses resampling to specify number of null permutations under the null to assess significance levels.  See if significance persist after increas resamp from 100 to 1000, which is the default.
# The x-intercept in the results is the distance at which objects are no more similar than expected by chance alone across the region.
# Run on CPL as that is giving correlation greater than 1, which according to Fletcher and Fortin might be due to small number of pairs of points used in that first distance bin or to outliers.  Perhaps, cluster of high Delaware sites (n visit 1 = 5) causes the correlation to be greater than 1.
# Looks like 71 pairs of points being used.
# With a lag, or bin distance, of 50 km only go out 10 bins, 500 km, and do a Bonferroni adjustment of 0.05/10 = 0.005
names(dg_1)
dg_cplvisit1 <- dg_1 %>% filter(wsa9 == "CPL" & visit_no == 1)
# N.B. default is resamp = 1000 so that can take 10 minutes for n =1090 sites
class(dg_cplvisit1)

correlo_n2o_ncf <- ncf::correlog(x = dg_cplvisit1$map_lon_dd, y = dg_cplvisit1$map_lat_dd, z = dg_cplvisit1$log_n2o, increment = 50, resamp = 1000, latlon = TRUE)
names(correlo_n2o_ncf)
plot(correlo_n2o_ncf) # works when pgmiress commented out
rm(dg_napvisit1, correlo_n2o_ncf)

# B/c returns a list of list have to access with [[]]
print(correlo_n2o_ncf[["n"]])
print(correlo_n2o_ncf[["mean.of.class"]])
print(correlo_n2o_ncf[["correlation"]])
head(correlo_n2o_ncf[["mean.of.class"]])
plot(correlo_n2o_ncf[["mean.of.class"]], correlo_n2o_ncf[["correlation"]])

class(correlo_n2o_ncf)
ncf::plot.correlog(correlo_n2o_ncf)

# CPL outlier check by working through all 10 pairwise removal of cluster of 5 high Delaware sites
out_cplvisit1 <- dg_1 %>% filter(wsa9 == "CPL" & visit_no == 1 & (site_id != "NLA17_DE-10003" & site_id != "NLA17_DE-10004"))

out_cplcorrelo_n2o <- ncf::correlog(x = out_cplvisit1$map_lon_dd, y = out_cplvisit1$map_lat_dd, z = out_cplvisit1$log_n2o, increment = 50, resamp = 1000, latlon = TRUE)

plot(out_cplcorrelo_n2o)
rm(out_cplvisit1, out_cplcorrelo_n2o)

# trying pgirmess package uses library spdep including moran.test and geary.test functions.  Distances are euclidean and in the same units as the spatial coordinates.  As run below, using geographical coordinates in decimal degrees so are distance classes retuned in decimal degrees?
# what if run on a spatial points data frame?
# Uses normality assumption, which is faster, to assess significance
coords_nap <- cbind(dg_napvisit1$map_lon_dd, dg_napvisit1$map_lat_dd)
correl_napn2o_pg <- pgirmess::correlog(coords_nap, dg_napvisit1$log_n2o, method = "Moran", nbclass = NULL, alternative = "two.sided")

head(round(correl_napn2o_pg, 2))
tail(round(correl_napn2o_pg, 2))
plot(correl_napn2o_pg)

# apply to SpatialPointsDataFrame that has projected coordinates and believe returning distance classes in meters
coords_sp_nap <- coordinates(nap_sp_visit1)
correl_spnapn2o_pg <- pgirmess::correlog(coords_sp_nap, nap_sp_visit1@data$log_n2o, method = "Moran", nbclass = 30, alternative = "two.sided")

head(round(correl_spnapn2o_pg, 2), 30)
tail(round(correl_spnapn2o_pg ,2))
plot(correl_spnapn2o_pg)
```

## Correlograms by Ecoregions
Doug Anderson wrote functions to deal with the list of list output from the correlog function in the ncf package.  He also wrote functions so that the correlogram of dissolved NO2 gas could be plotted by ecoregion using base plot, cowplot, or ggplot2 code.
```{r, correlogram_ecoregions}
# Original code from Doug Anderson at Neptune from file:  Neptune_correlogram_TDS05_Anderson.Rmd

# set the theme of the ggplot
theme_set(theme_bw()) # the theme I prefer is theme_bw()

## Can define a new function to plot
plot_correlog <- function(ncf_correlog, 
                          # names of list components
                          n_col = "n", 
                          mean_col = "mean.of.class",
                          cor_col = "correlation",
                          xint_col = "x.intercept",
                          pval_col = "p",
                          name_col = NULL, #can change to column that has names of interest
                          alpha = 0.05, # two sided significance level
                          # plot labels and adjustments
                          plot_title = "Correlogram", #can change to NULL
                          plot_subtitle = waiver(),
                          y_label = "Correlation",
                          x_label = "Distance (mean-of-class)",
                          pt_size = I(2), pt_color = "black",
                          ln_size = I(0.5), ln_color = "black", 
                          ln_type = "solid",
                          title_size = 14, subtitle_size = 14, 
                          axis_title_size = 14, axis_text_size = 14, 
                          title_color = "black", 
                          subtitle_color = "black",
                          axis_title_color = "black", 
                          axis_text_color = "black",
                          title_bold = TRUE, axis_bold = FALSE,
                          xaxis_color = "gray", xaxis_type = "solid",
                          theme_style = theme_bw()){
  # ncf_correlog is a list object exported from ncf::correlog() but can also be a dataframe with named columns

  # prepare a data set for plotting based on the list components or data frame columns
  dat <- data.frame(
    n = ncf_correlog[[n_col]],
    mean = ncf_correlog[[mean_col]],
    cor = ncf_correlog[[cor_col]],
    # xint = ncf_correlog[[xint_col]],
    pval = ncf_correlog[[pval_col]]
  )
  # create a variable to help plot the hollow and solid points
  # If a permutation test was performed, values significant at a nominal (two-sided) 5%-level 
  # will be represented by filled circles and non-significant values by open circles.
  dat$sig <- ifelse(dat$pval * 2 < alpha, TRUE, FALSE)

  ## Start the plot
  g <- ggplot(dat) +
    theme_style +
    geom_hline(yintercept = 0, color = xaxis_color, linetype = xaxis_type) +
    # geom_point(aes(x=mean, y=cor, shape=sig), size=pt_size, color=pt_color) +
    # geom_line(aes(x=mean, y=cor, group=1), size=ln_size, color=ln_color, linetype=ln_type) +
    lemon::geom_pointline(aes(x=mean, y=cor, shape=sig, group=1),
                          size = pt_size, color = pt_color, 
                          linetype = ln_type, linesize = ln_size, 
                          linecolor = ln_color) +
    scale_shape_manual(values=c("TRUE"=19, "FALSE"=1),
                       labels=c("TRUE"="Sig.", "FALSE"="Not Sig."),
                       guide=FALSE) +
    labs(x=x_label, y=y_label)

  ## add a title
  if(!is.null(name_col)){#include a title if column for name is provided or list object has a list item for name
    if(!is.null(plot_title)){ #if plot_title is a string, separate the plot_title with the other name by a ':'
      nm <- unique(as.character(ncf_correlog[[name_col]]))
      g <- g + 
        ggtitle(label = paste(plot_title, ": ", nm, sep=""), subtitle = plot_subtitle)
    } else{ #if plot_title is null, only include the name
      nm <- unique(as.character(ncf_correlog[[name_col]]))
      g <- g + 
        ggtitle(label = nm, subtitle = plot_subtitle)
    }
  } else{ # if there is no name column, only include the plot_title of interest (can be NULL)
    g <- g + 
      ggtitle(label = plot_title, subtitle = plot_subtitle) 
  }

  ## modify theme
  g <- g + 
    theme(plot.title = element_text(size=title_size, 
                                    hjust=0.5, #change to a value between 0 (left) and 1 (right)
                                    face=ifelse(title_bold == TRUE, "bold", "plain"), 
                                    color=title_color),
          plot.subtitle = element_text(size=subtitle_size, 
                                       hjust=0.5, 
                                       color=subtitle_color),
          axis.title = element_text(size=axis_title_size, 
                                    face=ifelse(axis_bold == TRUE, "bold", "plain"), 
                                    color=axis_title_color),
          axis.text = element_text(size=axis_text_size, #?theme
                                   color=axis_text_color), #?element_text
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
  return(g)
}

dg_1 <- select(dg,site.id:map.lon.dd, dissolved.n2o, WSA9, WSA9_NAME, ONOFFNET, nitrate.n.result,chla.result.vol, MAX.BF, temp.surf, temp.bottom, o2.surf, o2.bottom, lake.orgn, state, size_cat)
names(dg_1)
dg_1 <- clean_names(dg_1)
names(dg_1)
dg_1 <- dg_1 %>% mutate(log_n2o = round(log(dissolved_n2o),3))
## Subset dg_1 by visit_no == 1
dg_visit1 <- dg_1 %>% dplyr::filter(visit_no == 1)

## Perform the same test for each wsa9 using Z = "log_n2o"
## Start by defining a function
ncf_correlog <- function(dat, # data to send to ncf::correlog
                         z_col, # name of the column in the dataset for water quality parameter
                         i = 50, # increment to be used in ncf::correlog
                         N = 1000, # number of resamples in ncf::correlog
                         x_col = "map_lon_dd", # name of the column in the dataset for longitude
                         y_col = "map_lat_dd", # name of the column in the dataset for latitude
                         lat_lon = TRUE, # if the x and y coordinates for ncf::correlog are lat and lon
                         rem_missing = FALSE, # argument of ncf::correlog
                         run_quietly = FALSE){ # argument of ncf::correlog
  library(ncf)

  X = dat[[x_col]] #grabs the x_col from data.frame
  Y = dat[[y_col]] #grabs the y_col from data.frame
  Z = dat[[z_col]] #grabs the z_col from data.frame

  ## Use ncf::correlog
  crlg <- ncf::correlog(x = X, y = Y, z = Z, 
                        increment = i, resamp = N, latlon = lat_lon, 
                        na.rm = rem_missing, quiet = run_quietly)
  return(crlg)
}

## Split the dataframe by each wsa9: turns the data frame into a list 
##   where each list item is a data frame filtered to that wsa9
## lapply applies the function ncf_correlog on each list object (a data frame)
## the final product is a list of correlog objects
## this is similar to looping
## setting seed so get same result from permutations ncf::correlog function
set.seed(2035)
log_n2o_correlog_list <- dg_visit1 %>%
  split(.$wsa9) %>% # '.' means use the data frame that preceded this line
  lapply(function(x) ncf_correlog(dat = x, z_col = "log_n2o")) 
#ncf_correlog is the function name, 
#dat is the paramater that takes a data.frame,
#z_col is the parameter that takes a string name for a column of a water quality paramater

## look at the structure: list of 9 items where each item is a 
##   list containing 6 items: n, mean, correlation, x.int, p, call
str(log_n2o_correlog_list)

## Function to remove the list item 'call' which is of a different length 
##   than the other items (n, mean, correlation, x.int, and p)
## This function will also convert each list item to a dataframe
crlg_to_dat <- function(ncf_correlog, # ncf_correlog is a list object that was output from ncf::correlog
                        n_col = "n", 
                        mean_col = "mean.of.class",
                        cor_col = "correlation",
                        xint_col = "x.intercept",
                        pval_col = "p"){ 
  dat <- data.frame(
    n = ncf_correlog[[n_col]],
    mean = ncf_correlog[[mean_col]],
    cor = ncf_correlog[[cor_col]],
    xint = ncf_correlog[[xint_col]],
    pval = ncf_correlog[[pval_col]]
  )
  return(dat)
}

## log_n20_correlog_list is already a list so there is no need to split
log_n2o_correlog_dats <- lapply(log_n2o_correlog_list, 
                                function(x) crlg_to_dat(ncf_correlog = x))
str(log_n2o_correlog_dats) # each item in the list is now a dataframe

## combine into one dataframe to grab the wsa9 (notice the name no longer has an s at the end)

log_n2o_correlog_dat <- log_n2o_correlog_dats %>%
  dplyr::bind_rows(., .id = "wsa9")

str(log_n2o_correlog_dat)
head(log_n2o_correlog_dat)
## Make individual plots, the following object is a list containing a ggplot for each
log_n2o_correlog_plots <- log_n2o_correlog_dat %>%
  split(.$wsa9) %>% # splits the data frame by wsa9, which creates a list with 9 data frames
  lapply(function(x) plot_correlog(ncf_correlog = x, # use the data.frame for each wsa9 to make a plot
                                   plot_title = NULL,
                                   n_col = "n", 
                                   mean_col = "mean",
                                   cor_col = "cor",
                                   xint_col = "xint",
                                   pval_col = "pval",
                                   name_col = "wsa9",
                                   title_size = 14, 
                                   subtitle_size = 14, 
                                   axis_title_size = 14, 
                                   axis_text_size = 14))
## if hard to see writing run again but increase "title_size", "subtitle_size", "axis_title_size", "axis_text_size" as needed

## Look at each piece individually or scroll through
length(log_n2o_correlog_plots) # shows 9 items
# log_n20_correlog_plots
log_n2o_correlog_plots$CPL #or log_n20_correlog_plots[["CPL"]] or log_n2o_correlog_plots[[1]]
# log_n2o_correlog_plots$NAP
# log_n2o_correlog_plots$NPL
# log_n2o_correlog_plots$SAP
# log_n2o_correlog_plots$SPL
# log_n2o_correlog_plots$TPL
# log_n2o_correlog_plots$UMW
# log_n2o_correlog_plots$WMT
# log_n2o_correlog_plots$XER

## cowplot package (or other) can be used to group them into a matrix plot
log_n2o_correlog_plotgrid <- 
  cowplot::plot_grid(plotlist=log_n2o_correlog_plots, ncol=3)
print(log_n2o_correlog_plotgrid) # click 'Zoom' in plots quadrant, takes a few seconds to render
## can save the plot using 
# cowplot::ggsave(filename = "FILENAME.png", #saves a png version of the figure to the working folder
#                 plot = log_n20_correlog_plotgrid)
## can change .png to other forms
## can play with scale, width, and height to adjust the output of the figure saved to the png file

## alternatively, can make a facet plot since log_n20_correlog_dat is a dataframe in the correct format
names(log_n2o_correlog_dat) #"n"    "mean" "cor"  "xint" "pval" "wsa9"

## create column to shape point, sig == TRUE means two-sided test < 0.05
log_n2o_correlog_dat$sig <- ifelse(log_n2o_correlog_dat$pval * 2 < 0.05,
                                   TRUE, FALSE) 
head(log_n2o_correlog_dat)
## can change wsa9 to a factor to arrange what order the plots are shown, otherwise default arranges them alphabetically
# log_n20_correlog_dat$wsa9 <- factor(log_n20_correlog_dat, 
#                                     levels = c("CPL", "NAP", "NPL", "SAP", "SPL", "TPL", "UMW", "WMT", "XER"))

## make facet plot
log_n2o_correlog_facetplot <- ggplot(log_n2o_correlog_dat) +
  geom_hline(yintercept = 0, color = "gray", linetype = "solid") +
  # geom_point(aes(x=mean, y=cor, shape=sig), size=I(2), color="black") +
  # geom_line(aes(x=mean, y=cor, group=1), size=I(0.5), color="black", linetype="solid") +
  lemon::geom_pointline(aes(x=mean, y=cor, shape=sig, group=1), 
                        size=I(2), color = "black", 
                        linetype = "solid", linesize = I(0.5), 
                        linecolor = "black") +
  scale_shape_manual(values=c("TRUE"=19, "FALSE"=1), 
                     labels=c("TRUE"="Sig.", "FALSE"="Not Sig."),
                     guide=FALSE) +
  facet_wrap(~wsa9, ncol=3, scales="free") +
  labs(x="Distance (mean-of-class)", y="Correlation") +
  theme_bw() +
  theme(plot.title = element_text(size=14, hjust=0.5, 
                                  face="bold", color="black"),
        plot.subtitle = element_text(size=14, hjust=0.5, color="black"),
        axis.title = element_text(size=14, face="plain", color="black"),
        axis.text = element_text(size=14, color="black"),
        strip.text = element_text(size=14, color="black", face="bold"),
        strip.background = element_rect(fill="gray", color="black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
print(log_n2o_correlog_facetplot)
## can save the plot using 
# cowplot::ggsave(filename = "FILENAME2.png", #saves a png version of the figure to the working folder
#                 plot = log_n20_correlog_facetplot) 
## can change .png to other forms
## can play with scale, width, and height to adjust the output of the figure saved to the png file
